The quench action approach for integrable models: A Monte Carlo implementation

### 2015-09-24 draft/fs_qa_v1.tex
rev-24/09/2015

@@

The functional integral in~\eref{qa-d-ensemble} can be evaluated in the limit 
$L\to\infty$ using the saddle point approximation. One has to minimize the 
functional ${\mathcal F}(\pmb{\rho})$ defined as 
%
\begin{equation}
L{\mathcal F}(\pmb{\rho})\equiv 2\Re\log\langle\pmb{\rho}|\Psi_0
\rangle+S_{YY}(\pmb{\rho}(\lambda))  
\end{equation}
% 
with respect to $\pmb{\rho}$, under the constraint that the thermodynamic BGT 
equations~\eref{bgt-th} hold. Remarkably, for the quench with initial 
state the N\'eel state $|\Psi_0\rangle=|N\rangle$ the resulting saddle point root 
distributions $\rho_n^*(\lambda)$ can be obtained analytically~\cite{brockmann-2014}. 
The first few are given as 
%
\begin{eqnarray}
\label{rho_sp}
\fl\quad\rho^*_1(\lambda)=\frac{8(4+\lambda^2)}{\pi(19+3\lambda^2)(1+6\lambda^2+
\lambda^4)}\\
\fl\quad\rho_2^*(\lambda)=\frac{8\lambda^2(9+\lambda^2)(4+3\lambda^2)}{\pi(2+\lambda^2)
(16+14\lambda^2+\lambda^4)(256+132\lambda^2+9\lambda^4)}\\
\fl\quad\rho_3^*=\frac{8(1+\lambda^2)^2(5+\lambda^2)(16+\lambda^2)(21+\lambda^2)}
{\pi(19+3\lambda^2)(9+624\lambda^2+262\lambda^4+32\lambda^6+\lambda^8)
(509+5\lambda^2(26+\lambda^2))}.
\end{eqnarray}
%

@@

@@

This is obtained by usigng~\eref{neel-ov-th} and the saddle point root distributions 
$\rho_n^*$ (cf.~\eref{rho_sp} for the results up to $n=3$). Note that its position 
does not coincide with the peak of the overlap distribution function, as expected. 
This is reflecting the competition between the driving term and the entropy 
term in the quench action.

@@

Note that due to the locality of $Q_n$, the translational invariance of the initial 
state, and the periodic boundary conditions, the charge density $Q_n^{(0)}/L$ does 
not depend on the chain size. For the N\'eel state, in the thermodynamic limit 
$Q_n^{(0)}$ is obtained using~\eref{q0-th} and the quench action root distributions 
$\pmb{\rho}^*$ (cf.~\eref{rho_sp}).

@@

The BGT root distributions corresponding to the quench action steady state 
(cf.~\eref{rho_sp}) $\pmb{\rho}=\{\rho_n(\lambda)\}_{n=1}^{\infty}$ can be extracted 
from the Monte Carlo simulation.  This has been shown in Ref.~\cite{alba-2015} for 
the GGE representative state.

@@

This is demonstrated numerically in Figure~\ref{fig4:neel-rho} for the first two root 
distributions $\rho^*_1(\lambda)$ (panel (a)(b)) and $\rho^*_2(\lambda)$ (panel 
(c)(d)), plotted agains 
the BGT root $\lambda$. The data are the histograms of the BGT roots sampled 
in the Monte Carlo. The data are for a chain with $L=56$ sites and 
$N_{mcs}\sim 10^7$ Monte Carlo steps. All the data are divided by a factor 
$10^6$ for plotting convenience. The width of the histogram bins $\Delta\lambda$ 
is $\Delta\lambda\approx0.02$ and $\Delta\lambda\approx0.001$ for $\rho_1(\lambda)$ 
and $\rho_2(\lambda)$, respectively.

The histogram fluctuations are due to $N_{mcs}$ being finite, and are expected 
to vanish in the limit $N_{mcs}\to\infty$.

The continuous lines is the expected analytic result in the thermodynamic limit.

Clearly, $\rho_1(\lambda)$ is in good agreement with the Monte Carlo data in the 
whole range $-2\le\lambda\le2$ considered. 
On the other hand, larger deviations from the theoretical result are present 
for $\rho_2(\lambda)$. These deviations are larger on the tails of the 
distribution. This is expected since large $\lambda$ correspond to large 
quasimomenta, which are more sensitive to the finite size of the chain.

@@

@@

@@

>>>>

It is useful to rewrite~\eref{d-ensemble} as 
%
\begin{equation}
\label{qa-prob}
\langle{\mathcal O}\rangle=\sum\limits_{\alpha}\rho^{DE}(\alpha)
\langle\alpha|{\mathcal O}|\alpha\rangle,\quad\textrm{with}\quad 
\rho^{DE}(\alpha)=\exp(-2\Re{\mathcal E}(\alpha)).
\end{equation}
%
Here $\rho^{DE}(\alpha)$ is a diagonal ensemble density matrix, and ${\mathcal E}(\alpha)
\equiv-\log\langle\alpha|\Psi_0\rangle$, with $\Re$ denoting the real part.

@@

The functional integral in~\eref{qa-d-ensemble} can be evaluated in the limit 
$L\to\infty$ using the saddle point approximation. One has to minimize the 
functional ${\mathcal F}(\pmb{\rho})$ defined as 
%
\begin{equation}
L{\mathcal F}(\pmb{\rho})\equiv 2\Re\log\langle\pmb{\rho}|\Psi_0
\rangle+S_{YY}(\pmb{\rho}(\lambda))  
\end{equation}
% 
with respect to $\pmb{\rho}$, under the constraint that the thermodynamic BGT 
equations~\eref{bgt-th} hold. Remarkably, for the quench with initial 
state the N\'eel state $|\Psi_0\rangle=|N\rangle$ the resulting saddle point root 
distributions $\rho_n^*(\lambda)$ can be obtained analytically~\cite{brockmann-2014}. 
The first few are given as 
%
\begin{eqnarray}
\label{rho1-sp}
\fl\quad\rho^*_1(\lambda)=\frac{8(4+\lambda^2)}{\pi(19+3\lambda^2)(1+6\lambda^2+
\lambda^4)}\\\label{rho2-sp}
\fl\quad\rho_2^*(\lambda)=\frac{8\lambda^2(9+\lambda^2)(4+3\lambda^2)}{\pi(2+\lambda^2)
(16+14\lambda^2+\lambda^4)(256+132\lambda^2+9\lambda^4)}\\\label{rho3-sp}
\fl\quad\rho_3^*=\frac{8(1+\lambda^2)^2(5+\lambda^2)(16+\lambda^2)(21+\lambda^2)}
{\pi(19+3\lambda^2)(9+624\lambda^2+262\lambda^4+32\lambda^6+\lambda^8)
(509+5\lambda^2(26+\lambda^2))}.
\end{eqnarray}
%

@@

Interestingly, the data suggest the behavior
%
\begin{equation}
|\langle\lambda|N\rangle|^2\propto\exp(-\kappa L),
\end{equation}
%
with $0.18\lesssim\kappa\lesssim0.7$. The vertical dash-dotted line in 
Figure~\ref{fig0:neel-ov} is the overlap between the ground state of the 
$XXX$ chain and the N\'eel state. This is obtained using the ground state 
root distribution $\rho_1(\lambda)\propto1/\cosh(\pi\lambda)$  
and~\eref{neel-ov-th}. The dashed line denotes the N\'eel overlap 
$\sim 2/B(L,L/2)$ with the $S_z=0$ component of the ferromagnetic state.

@@

This is obtained by usigng~\eref{neel-ov-th} and the saddle point root distributions 
$\rho_n^*$ (cf.~\eref{rho1-sp}-\eref{rho3-sp} for the results up to $n=3$). Note that 
its position does not coincide with the peak of the overlap distribution function, as 
expected. This is reflecting the competition between the driving term and the entropy 
term in the quench action.

@@

Note that due to the locality of $Q_n$, the translational invariance of the initial 
state, and the periodic boundary conditions, the charge density $Q_n^{(0)}/L$ does 
not depend on the chain size. For the N\'eel state, in the thermodynamic limit 
$Q_n^{(0)}$ is obtained using~\eref{q0-th} and the quench action root distributions 
$\pmb{\rho}^*$ (cf.~\eref{rho1-sp}-\eref{rho3-sp}).

@@

The BGT root distributions corresponding to the quench action steady state 
(cf.~\eref{rho1-sp}-\eref{rho3-sp}) $\pmb{\rho}=\{\rho_n(\lambda)\}_{n=1}^{\infty}$ can be extracted 
from the Monte Carlo simulation.  This has been shown in Ref.~\cite{alba-2015} for 
the GGE representative state.

@@

This is demonstrated numerically in Figure~\ref{fig4:neel-rho} for the first two root 
distributions $\rho^*_1(\lambda)$ (panel (a)(b)) and $\rho^*_2(\lambda)$ (panel 
(c)(d)). The data are obtained using the algorithm described in subsection~\ref{sec:6.1} 
for chains with up to $L\le 56$ sites. The data for $L=48$ correspond to a Monte Carlo 
history with $N_{mcs}\sim 10^7$ Monte Carlo steps (cf.~\eref{qamc-obs}), whereas we 
used $N_{mcs}\sim 10^8$ for $L=56$.

Panel (a) and (c) show the  histograms of the $1$-string and $2$-string BGT roots sampled 
in the Monte Carlo, respectively for the chain with $L=56$ sites. The $y$-axis is 
divided by a factor $10^6$ for convenience. The width of the histogram bins $\Delta\lambda$ 
is $\Delta\lambda\approx0.02$ and $\Delta\lambda\approx0.001$ for $\rho_1(\lambda)$ 
and $\rho_2(\lambda)$, respectively. The histogram fluctuations are due both to the finite 
statistics (finite $N_{mcs}$) and to the finite size of the chain.

The extracted quench action saddle point root distributions $\rho^*_1(\lambda)$ and 
$\rho^*(\lambda)$ are shown in panels (b) and (d). The data are the same as in panel 
(a)(c). The normalization of the histograms is chosen such as to match the analytical 
results from~\eref{rho1-sp} and~\eref{rho2-sp}, i.e., $\int d\lambda
\rho^*_1(\lambda)\approx0.31$ and $\int d\lambda\rho^*_2(\lambda)\approx
0.015$. The Monte Carlo error bars shown in the Figure are obtained with a standard 
jackknife analysis~\cite{quenouille-1949,wolff-2004}. 
The continuous lines are the expected analytic result in the thermodynamic limit given 
in~\eref{rho1-sp} and~\eref{rho2-sp}. 
Clearly, $\rho_1(\lambda)$ is in good agreement with the Monte Carlo data in the 
whole range $-2\le\lambda\le2$ considered. The statistical error bars are smaller 
than the symbol size. The oscillations are due to the finite size of the chain, and 
decrease with the chain size (see the data for $L=48$). 
On the other hand, larger (oscillating) finite size effects are observed for $\rho_2(\lambda)$. 
These deviations are larger on the tails of the root distribution. This is expected since 
large $\lambda$ correspond to large quasimomenta, which are more sensitive to the finite 
size of the chain. Moreover, the Monte Carlo error bars are clearly visible. This is 
due to the fact that since $\int d\lambda\rho^*_2(\lambda)/\sum_n\int d\lambda
\rho_n^*(\lambda)\approx 0.04$, the statistics available to estimate $\rho_2^*(\lambda)$ 
is effectively reduced as compared to $\rho_1^*(\lambda)$. 
Finally, we numerically checked that this effect is more dramatic for the longer strings 
root distribution, and larger statistics and larger chain would be needed already  
to extract $\rho_3(\lambda)$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{conclusions}

We presented a Monte Carlo implementation of the quench action method for integrable 
spin chains. We focused on the spin-$1/2$ isotropic Heisenberg spin chain ($XXX$ chain) 
and the quench from the zero-momentum N\'eel state. 
The approach extends the Monte Carlo approach developed in Ref.~\cite{alba-2015} to 
simulate the Generalized Gibbs Ensemble in integrable models. The key idea of the 
approach is the Monte Carlo sampling of the model Hilbert space with the quench 
action, equivalently the Diagonal Ensemble, probability distribution given 
in~\eref{qa-prob}.

The approach is based on the knowledge of the 
overlaps between the pre-quench N\'eel state $|N\rangle$ and the eigenstates 
of the model that have been obtained recently~\cite{pozsgay-2014,brockmann-2014,
brockmann-2014a,brockmann-2014b,brockmann-2014c,piroli-2014}.

Besides the overlaps the method is based on the knowledge of the Hilbert space 
structure provided by the Bethe ansatz formalism and on the Bethe-Gaudin-Takahashi 
(BGT) equations. The approach is devised for finite-size systems. Thermodynamic 
quantities can be extracted using finite-size scaling.

We first focused on the importance of the so-called zero-momentum strings. 
These corresponds to components of the Bethe wavefunctions corresponding to 
bound states of many down spins (strings in the Bethe ansatz language) with 
zero center of mass momentum. These are usually neglected in the quench action 
approach~\cite{brockmann-2014}. The reason is that zero momentum strings 
leads to singularities in the overlap formulas, which make extracting the 
thermodynamic limit tricky. Moreover, it is argued that zero-momentum 
strings give rise only to scaling corrections to the quench action results, 
which are unimportant in the thermodynamic limit.

Importantly, one should observe that for a finite chain the vast majority of 
eigenstates contain zero-momentum strings. More precisely, for the quench from 
the N\'eel state the fraction of eigenstates with no zero-momentum strings 
vanish as $\propto L^{-1/2}$.

Moreover, the only eigenstates having non zero N\'eel overlaps (and non zero 
overlap with the Majumdhar-Ghosh state) are the so-called parity invariant 
eigenstates. The total number of parity-invariant eigenstates of the Heisenberg 
spin chain is given in terms of the chain length by a simple combinatorial 
formula that we provide. We provide similar combinatorial formulas for the 
total number of parity-invariant eigenstates at fixed number of down spins 
and fixed string content, i.e., the number of many-spin bound states present in 
each eigenstates.

In order to understand the importance of the zero-momentum strings we calculated 
the overlap distribution function including all the eigenstates of the chain for 
chains up to $L=38$ sites. We excluded the zero-momentum strings.

We focused on some exact sum rules for the local integrals of motion of the 
$XXX$ chain. Strikingly, the exclusion of the zero momentum strings gives rise 
to violations of the sum rules. Precisely, we numerically observe that the 
sum rules exhibit the vanishing behavior $\propto L^{-\frac{1}{2}}$, which 
reflects the vanishing of the fraction of eigenstates with no zero-momentum 
strings.

The behavior $\propto L^{-\frac{1}{2}}$ is not generic, but it depends on the 
pre-quench initial state. We demonstrated that by considering the quench from 
the Majumdar-Ghosh state. We observed that the sum rule vanish as $1/L$, again 
reflecting the vanishing as $1/L$ of the fraction of eigenstates with no 
zero-momentum strings.

We observed that the vast majority of the eigenstates have neglegible overlap 
with the N\'eel state. This implies that the eigenstates relevant for the 
quench action are a small fraction of the total number of eingestates having 
in principle non-zero N\'eel overlap.

Next we presented a Monte Carlo implementation of the quench action method. 
This is obtained by resampling with the quench action probability measure 
the Hilbert space of the $XXX$ chain, restricted to the sector of the  Hilbert 
space with no zero-momentum strings.

This is obtained using a standard Metropolis algorithm. Importantly, we 
numerically demonstrated that the only consequence of the Hilbert space 
restriction to strings with finite momentum are scaling corrections $1/L$ 
to physical observables.

Specifically, we focused on the sum rules for the local conserved quantities 
of the $XXX$ chain. We numerically observed that, although for finite chains 
violations of the sum rules are present, these violations decay upon increasing 
the chain size, and the sum rules are restored in the thermodynamic limit.

Finally, we extracted from the Monte Carlo the BGT root distributions characterizing 
the quench action saddle point. These are obtained from the BGT roots identifying 
the $XXX$ chain eigenstates sampled in the Monte Carlo. We observed that, although 
the finite size gives rise to oscillating scaling corrections, already for 
relatively small chains with $L=56$ sites, the first BGT root distribution 
$\rho^*_1(\lambda)$ is in impressive agreement with the analytic result in 
the thermodynamic limit.

@@

\bibitem{bloch-2008}
I.~Bloch, J.~Dalibard, and W.~Zwerger, Rev.\ Mod.\ Phys.\ {\bf 80},
885 (2008).

\bibitem{greiner-2002}
M.~Greiner, O.~Mandel, T. H\"ansch, and I.~Bloch, Nature (London)
{\bf 419}, 51 (2002).

\bibitem{kinoshita-2006}
T.~Kinoshita, T.~Wenger, and D.~S.~Weiss, Nature (London) {\bf 440},
900 (2008).

\bibitem{hofferberth-2007}
S.~Hofferberth, I.~Lesanovsky, B.~Fischer, T.~Schumm, and J.~Schiedmayer,
Nature (London) {\bf 449}, 324 (2007).

\bibitem{trotzky-2012}
S.~Trotzky, Y.-A.~Chen, A.~Flesch, I.~P.~McCulloch, U.~Schollw\"ock,
J.~Eisert, and I.~Bloch, Nature Phys.\ {\bf 8}, 325 (2012).

\bibitem{gring-2012}
M.~Gring, M.~Kuhnert, T.~Langen, T.~Kitagawa, B.~Rauer, M.~Schreitl,
I.~Mazets, D.~A.~Smith, E.~Demler, and J.~Schmiedmayer, Science {\bf 337},
6100 (2012).

\bibitem{cheneau-2012}
M.~Cheneau, P.~Barmettler, D.~Poletti, M.~Endres, P.~Schaua, T.~Fukuhara,
C.~Gross, I.~Bloch, C.~Kollath, and S.~Kuhr, Nature (London) {\bf 481},
484 (2012).

\bibitem{schneider-2012}
U.~Schneider, L.~Hackeruller, J.~P.~Ronzheimer, S.~Will, S.~Braun, T.~Best,
I.~Bloch, E.~Demler, S.~Mandt, D.~Rasch, and A.~Rosch, Nature\ Phys.\
{\bf 8}, 213 (2012).

\bibitem{kunhert-2013}
M.~Kuhnert, R.~Geiger, T.~Langen, M.~Gring, B.~Rauer,
T.~Kitagawa, E.~Demler, D.~Adu Smith, and J.~Schmiedmayer, Phys.\ Rev.\
Lett.\ {\bf 110}, 090405 (2013).

\bibitem{langen-2013}
T.~Langen, R.~Geiger, M.~Kuhnert, B.~Rauer, and J.~Schmiedmayer,
Nature\ Phys.\ {\bf 9}, 640 (2013).

\bibitem{meinert-2013}
F.~Meinert, M.~J.~Mark, E.~Kirilov, K.~Lauber, P.~Weinmann,
A.~J.~Daley, and H.-C.~Nagerl, Phys.\ Rev.\ Lett.\ {\bf 111},
053003 (2013).

\bibitem{fukuhara-2013}
T.~Fukuhara, A.~Kantian, M.~Endres, M.~Cheneau, P.~Schaua, S.~Hild, C.~Gross,
U.~Schollw\"ock, T.~Giamarchi, I.~Bloch, and S.~Kuhr, Nature\ Phys.\ {\bf 9},
235 (2013).

\bibitem{ronzheimer-2013}
J.~P.~Ronzheimer, M.~Schreiber, S.~Braun, S.~S.~Hodgman, S.~Langer, I.~P.~McCulloch,
F. Heidrich-Meisner, I.~Bloch, and U.~Schneider, Phys.\ Rev.\ Lett.\ {\bf 110},
205301 (2013).

\bibitem{braun-2014}
S.~Braun, M.~Friesdorf, S.~Hodgman, M.~Schreiber, J.~Ronzheimer, A.~Riera, M.~del Rey,
I.~Bloch, J.~Eisert, and U.~Schneider, PNAS {\bf 112}, 3641 (2015).

\bibitem{langen-2015}
T.~Langen, S.~Erne, R.~Geiger, B.~Rauer, T.~Schweigier, M.~Kuhnert, W.~Rohringer,
I.~E.~Mazets, T.~Gasenzer, J.~Schmiedmayer, Science {\bf 348}, 6231 (2015).

\bibitem{polkovnikov-2011}
A.~Polkovnikov, K.~Sengupta, A~Silva, and M.~Vengalattore, Rev.\ Mod.\ Phys.\
{\bf 83}, 863 (2011).

@@

\bibitem{brockmann-2014c}
M.~Brockmann, J.~De~Nardis, B.~Wouters, and J.-S.~Caux, J.\ Phys.\ A:\ 
Math.\ Theor. {\bf 47}, 145003 (2014).

\bibitem{brockmann-2014b}
M.~Brockmann, J.\ Stat.\ Mech.\ (2014) P05006.

@@

\bibitem{quenouille-1949} 
M.~H.~Quenouille, Ann.\ Math.\ Statist.\ {\bf 20}, 355 (1949).

\bibitem{wolff-2004} 
U.~Wolff, Comput.\ Phys.\ Comm.\ {\bf 156}, 143 (2004).

\bibitem{piroli-2014}
L.~Piroli and P.~Calabrese, J.\ Phys.\ A {\bf 47}, 385003 (2014).
